---
title: "Chapter 10 Part 2"
author: "Sean McMannamy"
date: "10/1/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(factoextra)
library(cluster)
library(fpc)
library(dplyr)
```

## Data Set
Use the diamonds data set within the ggplot2/tidyverse package and perform k-means, k-mediods, CLARA, BIRCH, CHAMELEON, and DBSCAN clustering. Do not use the price variable in your clustering. Once you have completed each clustering determine the mean and standard deviation of the price for each cluster. You must then determine which technique was the most useful for this situation and why that is.

```{r}
df <- diamonds
```

## Determining K

```{r}
head(df)
df_e <- df[,-7]
df_e$cut <- as.numeric(df$cut)
df_e$color <- as.numeric(df$color)
df_e$clarity <- as.numeric(df$clarity)

set.seed(55)
k.max <- 10
wss <- sapply(1:k.max,
              function(k) (kmeans(df_e, k, nstart=50, iter.max = 15)$tot.withinss))
plot(1:k.max, wss, xlab="clusters", ylab="wss")
df_e <- scale(df_e)
```

## K means

```{r}
k3 <- kmeans(df_e, centers = 3, nstart = 25)

df$cluster <- k3$cluster
mean_price <- df %>%
  group_by(cluster) %>%
  summarise(mean(price))
mean_price
sd_price <- df %>%
  group_by(cluster) %>%
  summarise(sd(price))
sd_price
# fviz_cluster(k3, data = df)
```

## K Medoids

```{r}
kmed <- pam(df_e, k = 3)

df$cluster <- kmed
mean_price <- df %>%
  group_by(cluster) %>%
  summarise(mean(price))
mean_price
sd_price <- df %>%
  group_by(cluster) %>%
  summarise(sd(price))
sd_price
# fviz_cluster(kmed, data = df)
```

## CLARA

```{r}
clara <- clara(df_e, 3, metric = "euclidean", stand = FALSE, 
               samples = 5, pamLike = FALSE)

df$cluster <- clara
mean_price <- df %>%
  group_by(cluster) %>%
  summarise(mean(price))
mean_price
sd_price <- df %>%
  group_by(cluster) %>%
  summarise(sd(price))
sd_price
```

## DBSCAN

```{r}
dbscan <- dbscan(df_e, eps = 0.45, MinPts = 5)

df$cluster <- dbscan
mean_price <- df %>%
  group_by(cluster) %>%
  summarise(mean(price))
mean_price
sd_price <- df %>%
  group_by(cluster) %>%
  summarise(sd(price))
sd_price
```

I believe using DBSCAN is the most useful technique in this situation because the data points are mostly in a highly concentrated part of the plot together. There is a high level of density. This algorithm is able to deal with this appropriately and still provide a useful clustering of the data.
